---
title: "BB503/BB602 - R Training - Week XII"
author: "Ege Ulgen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Logistic Regression

The data we'll use is `birthwt` from the `MASS` package. The `birthwt` data frame has 189 rows and 10 columns. The data were collected at Baystate Medical Center, Springfield, Mass during 1986.

```{r data}
# install.packages("MASS")
library(MASS)

data(birthwt) 
?birthwt

dim(birthwt)

head(birthwt)

# turn categorical variables into factor
birthwt$low <- as.factor(birthwt$low)
birthwt$race <- as.factor(birthwt$race)
birthwt$smoke <- as.factor(birthwt$smoke)
birthwt$ht <- as.factor(birthwt$ht)
birthwt$ui <- as.factor(birthwt$ui)

summary(birthwt)
```

We'll be using logistic regression to identify risk factors associated with low infant birth weight (birth weight less than 2.5 kg).

```{r lr1}
fit0 <- glm(low~.-bwt, data = birthwt, family = binomial)
summary(fit0)
```

We'll use only the significant variables:

```{r lr2}
fit1 <- glm(low~lwt + race + smoke + ht, data = birthwt, family = binomial)
summary(fit1)
```

The final model:

```{r lr3}
fit_final <- glm(low~I(lwt - min(lwt)) + race + smoke + ht, data = birthwt, family = binomial)
summary(fit_final)
```

```{r lr_coef}
coef(fit_final)

# OR
exp(coef(fit_final))[-1]

# % change in odds
(exp(coef(fit_final)) - 1)[-1] * 100 
```

```{r lr_vis}
# install.packages("sjPlot")
sjPlot::plot_model(fit_final)
```

# Poisson Regression

The data we'll use is `epilepsy` from the `HSAUR` package. The dataset is for a randomized clinical trial investigating the effect of an anti-epileptic drug (Progabide).

```{r poi_data}
# install.packages("HSAUR")
library(HSAUR)
?epilepsy

data("epilepsy")

dim(epilepsy)
head(epilepsy)

summary(epilepsy)
```

We'll only inspect period 4:

```{r subset}
epilepsy4 <- epilepsy[epilepsy$period == 4, ]

boxplot(epilepsy4$seizure.rate~epilepsy4$treatment)
```

Let's inspect the effect of treatment adjusting for `base` and `age`:

```{r pois_model}
fit_pois <- glm(seizure.rate ~ treatment + I(base - min(base)) + I(age - min(age)), data = epilepsy4, family = poisson)
summary(fit_pois)
```

```{r pois_interpret}
(exp(coef(fit_pois)[-1]) - 1) * 100
```

```{r pois_vis}
sjPlot::plot_model(fit_pois)
```
