---
title: "BB503/BB602 - R Training - Week II"
output: pdf_document
---

# Introduction

This week, we will be working with the AIDS data set[^1]. Let's start by reading the data. Notice that we had to specify that the separator (via the `sep` argument) is spaces (instead of the default of `read.delim()` which is tabs).

[^1]: Source: Applied longitudinal analysis [Internet]. [cited 2021 Sep 27]. Available from: <https://content.sph.harvard.edu/fitzmaur/ala2e/>

```{r read}
aids_df <- read.delim("../data/aids_dataset.txt", sep = " ")
```

For any function in R, we can use `?function_name` to view the help page for that function:

```{r help}
?read.delim
```

We can view the first 6 (by default) lines using the function `head()` and, similarly, the last lines using the `tail()` function:

```{r head_tail}
head(aids_df)

# print first 2 lines
head(aids_df, 2)

tail(aids_df)
```

Using `str()`, we can display the structure of our data frame `aids_df`. For later use, we turn the variables `treatment` and `gender` into factors.

```{r str}
str(aids_df)
aids_df$treatment <- as.factor(aids_df$treatment)
aids_df$gender <- as.factor(aids_df$gender)
str(aids_df)
```

# Examining Distributions

## Mean/Median/Mode

We can inspect the mean of a vector using the function `mean()`:

```{r mean}
# mean of age
mean(aids_df$age)

# mean of CD4 count at week 1
mean(aids_df$cd4_1)
```

Similarly, we can calculate the median of a vector using the function `median()` :

```{r median}
# median of CD4 count at week 2
median(aids_df$cd4_2)
```

Mode can be calculated using the function `table()` (calculates the frequency of each value):

```{r mode}
tbl <- table(aids_df$cd4_1)
sort(tbl, decreasing = TRUE)
```

## Quantiles

For calculating quantiles given a variable (a vector), we use the `quantile()` function:

```{r quantile}
# 10th, 25th, 33rd and 78th percentiles of age
quantile(aids_df$age, probs = c(0.1, 0.25, 0.33, 0.78))
```

By default, R uses "type 7" out of the 9 available algorithms for calculating quantiles. We can change this by altering the `type` argument:

```{r quantile2}
# 10th, 25th, 33rd and 78th percentiles of age - type 2
quantile(aids_df$age, probs = c(0.1, 0.25, 0.33, 0.78), type = 2)
```

## Spread

### Range

```{r range}
# range of CD4 counts at week 1
max(aids_df$cd4_1) - min(aids_df$cd4_1)
range(aids_df$cd4_1)
```

### Inter-quantile range

```{r IQR}
### IQR of CD4 counts at week 1
# calculate Q1 and Q3
quantile(aids_df$cd4_1, probs = c(0.25, 0.75))

# IQR = Q3 - Q1
36 - 11
```

alternatively:

```{r IQR2}
IQR(aids_df$cd4_1)
IQR(aids_df$cd4_1, type = 2)
```

### Variance and standard deviation

```{r var_sd}
# Variance of CD4 counts at week 1
var(aids_df$cd4_1)

# Std. deviation of CD4 counts at week 1
sqrt(var(aids_df$cd4_1))
sd(aids_df$cd4_1)
```

## Outliers

We can manually determine outliers in a variable using the definition provided in the slides.

```{r outliers}
### Outliers for CD4 counts at weeek 1
Q1 <- quantile(aids_df$cd4_1, 0.25)
Q3 <- quantile(aids_df$cd4_1, 0.75)

IQR_val <- IQR(aids_df$cd4_1)

UL <- Q3 + 1.5 * IQR_val
LL <- Q1 - 1.5 * IQR_val

cond <- aids_df$cd4_1 > UL | aids_df$cd4_1 < LL
table(cond)
which(cond)
aids_df$cd4_1[cond]
```

We can visualize the boxplot of this variable to see the outlier values:

```{r boxplot}
boxplot(aids_df$cd4_1)
boxplot(aids_df$cd4_1, horizontal = TRUE)
```

Using the function `boxplot.stats()`, to gather and display the statistics necessary for producing box plots, including outliers.

```{r boxplot_stats}
boxplot.stats(aids_df$cd4_1)
```

## The `summary()` function

```{r summary}
summary(aids_df$cd4_1)

summary(aids_df)
```

## The "improved" summary function

```{r new_summary}
new_summary <- function(x){ 
    out <- list(
        min = min(x),
        max = max(x),
        quants = quantile(x, prob = c(0.1, 0.25, 0.5, 0.75, 0.9),
                          type = 2),
        mean    = mean(x),
        var     = var(x),
        std_dev = sd(x),
        length  = length(x)
    )
  class(out) <- "new_summary"
  return(out) 
}

print.new_summary <- function(object, ...){
    cat("Min:", object$min, "\n")
    cat("10th percentile:", object$quants[1], "\n")
    cat("25th percentile:", object$quants[2], "\n")
    cat("50th percentile (median):", object$quants[3], "\n")
    cat("Mean:", object$mean, "\n")
    cat("75th percentile:", object$quants[4], "\n")
    cat("90th percentile:", object$quants[5], "\n")
    cat("Max:", object$max, "\n")
    cat("Var:", object$var, "\n")
    cat("Sd:", object$std_dev, "\n")
    cat("Length:", object$length, "\n")
}
```

Examples:

```{r new_summary2}
new_summary(aids_df$cd4_1)
new_summary(aids_df[aids_df$treatment == "trt1", "cd4_1"])
new_summary(aids_df[aids_df$treatment == "trt2", "cd4_1"])
new_summary(aids_df[aids_df$treatment == "trt3", "cd4_1"])
new_summary(aids_df[aids_df$treatment == "trt4", "cd4_1"])
```

# Examining Relationships

## Scatter plots

Age vs. baseline (week 1) CD4 cell counts:

```{r scattter}
plot(aids_df$age, aids_df$cd4_1)
plot(cd4_1 ~ age, aids_df)
```

## Correlation

### Pearson Correlation

```{r pearson}
?cor
cor(aids_df$age, aids_df$cd4_1)
```

### Spearman's Rank Correlation

```{r spearman}
cor(aids_df$age, aids_df$cd4_1, method = "spearman")
```

### Correlation Test

```{r cor_test}
cor.test(aids_df$age, aids_df$cd4_1, method = "spearman")
```
