---
title: "BB503/BB602 - R Training - Week III"
author: "Ege Ulgen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R tips

-   If you can't figure out how to solve an issue, Google is your friend. e.g., "how to calculate mode r"

-   If you need help with the usage of a function, type `?function_name`. e.g. `?quantile`

-   If you get an error, and cannot fix it. C/P the error into Google. Someone else most likely had a similar problem

-   Some resources for learning the basic syntax of R;

    -   Codecademy - <https://www.codecademy.com/learn/learn-r>

    -   RStudio Cloud Primers - <https://rstudio.cloud/learn/primers>

    -   Dataquest - <https://www.dataquest.io/course/introduction-to-data-analysis-in-r/>

    -   R for Data Science - <https://r4ds.had.co.nz/index.html>

-   Interesting read: <https://www.dataquest.io/blog/learn-r-for-data-science/>

-   Some resources you might want to follow:

    -   R-Bloggers - <https://www.r-bloggers.com/>

    -   STHDA - <http://www.sthda.com/english/>

    -   Quick-R - <https://www.statmethods.net/>

# R Basics

## Conditional Statements

```{r cond0}
### simple conditional (Boolean) statements
3 == 3
3 == 4
3 != 4
3 > 4
3 <= 4

5 %% 3 == 0
5 %% 3 == 2

### multiple conditions
# and
3 == 3 & 3 > 4
# or
3 == 3 | 3 > 4
# > 2 conditions
3 == 3 | (3 > 4 & 5 != 6)

!TRUE


```

We can use these conditions to selectively execute commands when these condition(s) are met:

```{r cond1}
# after a complex calculation X becomes 3
x <- 3

if (x == 3) {
    print("x is equal to 3")
    # other commands that should be executed if the statement is true go here
}
```

Using `else if` we can add even more conditions:

```{r cond2}
# after a complex calculation X becomes 5
x <- 5

if (x == 3) {
    print("x is equal to 3")
    # other commands that should be executed if this first statement is true go here
} else if (x > 3) {
    print("x is greater than 3")
    # other commands that should be executed if this second statement is true go here
}
```

Finally, using else we can execute any commands if none of the conditions are met:

```{r cond3}
# after a complex calculation X becomes 1
x <- 1

if (x == 3) {
    print("x is equal to 3")
    # other commands that should be executed if this first statement is true go here
} else if (x > 3) {
    print("x is greater than 3")
    # other commands that should be executed if this second statement is true go here
} else {
    print("x is less than 3")
    # other commands that should be executed if no condition is true go here
}
```

## Loops

We may wish to execute certain commands repeatedly until a certain condition is met. For this purpose, we can use a `while` loop:

```{r while}
# initialize
current_number <- 1
while(current_number <= 10) {
    # execute any commands
    print(paste("Current #:", current_number))
    # update
    current_number <- current_number + 1
}
```

We may wish to execute certain commands repeatedly for a set amount of times. For this purpose, we can use a `for` loop:

```{r for}
for (i in 1:10) {
    # execute any commands
    print(paste("Current #:", i))
}
```

## Putting it all together - The Fizz Buzz Game

Fizz buzz is a group word game for children to teach them about division. Players take turns to count incrementally,

-   replacing any number divisible by 3 with the word "fizz"

-   replacing any number divisible by 5 with the word "buzz"

```{r fizzbuzz}
for (i in 1:100) {
    if (i %% 3 == 0) {
        print("Fizz")
    } else if (i %% 5 == 0) {
        print("Buzz")
    } else {
        print(i)
    }
}
```

For a variation of the game, any number divisible by both 3 and 5 is replaced with the word "fizzbuzz"

```{r fizzbuzz2}
for (i in 1:100) {
    if (i %% 3 == 0 & i %% 5 == 0) {
        print("FizzBuzz")
    } else if (i %% 3 == 0) {
        print("Fizz")
    } else if (i %% 5 == 0) {
        print("Buzz")
    } else {
        print(i)
    }
}
```

## Functions

Functions are modules of code that accomplish a specific task. They take in data, process it, and return a result.

The exponentiation function $f(x,y)=x^y$ can be coded in R as:

```{r power}
pow <- function(x, y) {
    output <- x ^ y
    return(output)
}
```

E.g., $f(2,3)=2^3=8$:

```{r power2}
pow(2, 3)
```

Here,

-   `pow` is the name of the function, and it is arbitrary

-   `function(...)` is for defining any function in R

-   `x` and `y` are input names (arguments), and names are arbitrary

-   `output` is the object that stores the result (not created globally)

-   `return` is the function for outputting the result

We may define default values for the arguments:

```{r power3}
pow2 <- function(x, y = 2) {
    output <- x ^ y
    return(output)
}
pow2(3)
pow2(3, 3)
```

## Another Solution for Fizz-Buzz

```{r fizzbuzz_func}
fizz_buzz <- function(number) {
    if (number %% 3 == 0 & number %% 5 == 0) {
        return("FizzBuzz")
    }
    
    if (number %% 3 == 0) {
        return("Fizz")
    }
    
    if (number %% 5 == 0) {
        return("Buzz")
    }
    return(number)
}

fizz_buzz(2)
fizz_buzz(3)
fizz_buzz(5)
fizz_buzz(15)

sapply(1:100, fizz_buzz)
?sapply
```

## The `apply` family of functions

```{r apply}
mat <- matrix(rnorm(30), nrow = 5, ncol = 6)
sample_types <- c("Tumor", "Tumor", "Tumor", "Control", "Control")

?apply
# apply mean to each row
apply(mat, 1, mean)

# apply sd to each column
apply(mat, 2, sd)
```

```{r sapply}
# Apply a function over a list or vector
sapply(mat[1, ], function(x) x^3)
```

```{r tapply}
# Apply a function over a ragged array
# mean values of 1st variable per sample type
tapply(mat[, 1], sample_types, mean)
```

# Probability

## Basic Simulation - Flipping a fair coin

```{r simul1}
# by this you get the same answer all the time
set.seed(321)

# number of trials
n <- 5000

# tossing coins
results <- c() # to store results
average_at_i <- c() # to store averages (will approximate 0.5)
for (i in 1:n) {
    current_res <- sample(c(0, 1), 1)
    results <- c(results, current_res)
    average_at_i <- c(average_at_i, mean(results))
}

# plot resulting averages by each trial
plot(average_at_i, xlab = "Trial number", ylab = "Proportion of 1s")
plot(average_at_i, xlab = "Trial number", ylab = "Proportion of 1s", type = "l")
abline(h = 0.5, col = "red")
```
