---
title: "BB503/BB602 - R Training - Week VII"
author: "Ege Ulgen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# One-sample t-Test

The dataset we'll use contains information on 78 people using one of three diets.

```{r diet1}
diet_df <- read.csv("../data/Diet_R.csv")

head(diet_df)

# turn categorical variables into factor
diet_df$Diet <- as.factor(diet_df$Diet)
diet_df$gender <- as.factor(diet_df$gender)

# create new variable
diet_df$Weight_Change <- diet_df$weight6weeks - diet_df$pre.weight

summary(diet_df)
```

Let's look at the overall distribution of weight change and the distributions by Diet type:

```{r diet2}
# overall
hist(diet_df$Weight_Change)
boxplot(diet_df$Weight_Change)

# by diet type
boxplot(diet_df$Weight_Change~diet_df$Diet)
```

We can observe that `Weight_Change` seems to follow a normal distribution (we'll later learn how to test normality of a variable). Moreover, diet 3 seems to result in a higher decrease compared to diets 1 & 2. We'll test the difference between mean weight changes between diet 1 and 3 in the next section. For now, let's test whether the overall mean weight change is significantly different than -3.

## 1. Check assumptions, determine $H_0$ and $H_a$, choose $\alpha$

Inspecting the histogram of overall weight change, we concluded that it is normally-distributed.

$H_0: \mu = -3$ and $H_a: \mu \neq -3$

Let's choose $\alpha=0.05$

## 2. Calculate the appropriate test statistic

$$
t_H = \frac{\bar{X} - \mu}{s / \sqrt{n}}
$$

```{r t_stat}
t_stat <- (mean(diet_df$Weight_Change) - (-3)) / (sd(diet_df$Weight_Change) / sqrt(nrow(diet_df)))
t_stat

df <- nrow(diet_df) - 1
```

## 3. Calculate critical values/p value

### Critical values

```{r t_crit}
C1 <- qt(0.05/2, df = df)
C2 <- qt(1 - 0.05/2, df = df)
C1; C2

# t_stat in rejection zone?
t_stat < C1 | t_stat > C2
```

### p value

```{r t_p}
p_val <- 2 * (1 - pt(abs(t_stat), df = df))
p_val
```

### Confidence Interval

The 95% Confidence Interval for $\mu$:

$$
95\%\ CI = [\bar{X} + t_{\alpha/2} \frac{s}{\sqrt{n}},\ \bar{X} + t_{1- \alpha/2} \frac{s}{\sqrt{n}}]
$$

```{r t_CI}
SE <- sd(diet_df$Weight_Change) / sqrt(nrow(diet_df))

mean(diet_df$Weight_Change) + C1 * SE; mean(diet_df$Weight_Change) + C2 * SE
```

## 4. Decide whether to reject/fail to reject $H_0$

-   The calculated test statistic falls within the rejection region

-   p value \< $\alpha$

We reject the null hypothesis.

"With 95% confidence, there is enough evidence to say mean weight change is significantly different than 0."

"The overall mean weight change was found to be significantly different than -3 (t-test p \< 0.001, 95% CI = \[-4.42, -3.27\])"

### Using `t.test()`

```{r t_test_r}
?t.test
t.test(diet_df$Weight_Change, mu = -3)

# change mu
t.test(diet_df$Weight_Change, mu = -4)

# change alternative
t.test(diet_df$Weight_Change, mu = -3, alternative = "less")

# change conf. level
t.test(diet_df$Weight_Change, mu = -3, conf.level = .99)
```

# Two-sample t-Test

Let's test whether the mean weight changes of diet 1 and 3 are different.

```{r 2sample}

sub_df <- subset(diet_df, Diet %in% c(1, 3))

### compare variances (F-test)
var.test(sub_df$Weight_Change[sub_df$Diet == 1], sub_df$Weight_Change[sub_df$Diet == 3])
# more compactly
var.test(Weight_Change~Diet, data = sub_df)


res <- t.test(Weight_Change~Diet, data = sub_df, var.equal = TRUE)
res

res$p.value
res$conf.int
res$estimate
```

# Paired t-Test

A study tested whether cholesterol was reduced after using a certain brand of margarine as part of a low fat, low cholesterol diet. The subjects consumed on average 2.31g of the active ingredient, stanol easter, a day. This data set contains information on 18 people using margarine to reduce cholesterol over three time points.

```{r paired1}
chol_df <- read.csv("../data/Cholesterol_R.csv")
head(chol_df, 3)

# turn categorical variable into factor
chol_df$Margarine <- as.factor(chol_df$Margarine)

summary(chol_df)
```

For the overall data, let's compare whether there is a significant change between `Before` and `After8weeks`:

```{r paired2}
boxplot(chol_df$Before, chol_df$After8weeks)

# check normality
hist(chol_df$After8weeks - chol_df$Before)

t.test(chol_df$After8weeks, chol_df$Before, paired = TRUE)

t.test(chol_df$After8weeks - chol_df$Before)
```
