---
title: "BB503/BB602 - R Training - Week VIII"
author: "Ege Ulgen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of Variance (ANOVA)

We'll work on the diet data from last week. Remember there are 3 different diets:

```{r aov_data}
diet_df <- read.csv("../data/Diet_R.csv")

head(diet_df)

# turn categorical variables into factor
diet_df$Diet <- as.factor(diet_df$Diet)
diet_df$gender <- as.factor(diet_df$gender)

# create new variable
diet_df$Weight_Change <- diet_df$weight6weeks - diet_df$pre.weight

summary(diet_df)

boxplot(diet_df$Weight_Change~diet_df$Diet)
```

Let's test the hypothesis that at least one of the mean weight changes is different from the others between the three diets.

## 1. Check assumptions, determine $H_0$ and $H_a$, choose $\alpha$

Inspecting the histograms of weight change by diet, we conclude that they are normally-distributed:

```{r aov_norm}
par(mfrow = c(1, 3))
hist(diet_df$Weight_Change[diet_df$Diet == 1])
hist(diet_df$Weight_Change[diet_df$Diet == 2])
hist(diet_df$Weight_Change[diet_df$Diet == 3])
par(mfrow = c(1, 1))
```

$H_0: \mu_1 = \mu_2 = \mu_3$ and $H_a:$ at least one mean is different

Let's choose $\alpha=0.05$

## 2. Calculate the appropriate test statistic

### 2.1. Calculate the grand mean and means per group

```{r aov_mean}
grand_mean <- mean(diet_df$Weight_Change)
mean1 <- mean(diet_df$Weight_Change[diet_df$Diet == 1])
mean2 <- mean(diet_df$Weight_Change[diet_df$Diet == 2])
mean3 <- mean(diet_df$Weight_Change[diet_df$Diet == 3])
```

### 2.2. Calculate the (total, between group and within group) sum of squared error

Between group sum of squared error = $\sum n_i(\bar{X_i} - \bar{X})^2$

```{r aov_sse}

SS_bw <- sum(diet_df$Diet == 1) * (mean1 - grand_mean)^2 + 
    sum(diet_df$Diet == 2) * (mean2 - grand_mean)^2 +
    sum(diet_df$Diet == 3) * (mean3 - grand_mean)^2

SS_tot <- sum((diet_df$Weight_Change - grand_mean)^2) 

SS_wi <- SS_tot - SS_bw
```

### 2.3. Calculate degrees of freedom

```{r aov_df}
df_tot <- nrow(diet_df) - 1
df_bw <- 3 - 1
df_wi <- df_tot - df_bw
```

### 2.4. Calculate mean squared errors

```{r aov_mse}
MSE_bw <- SS_bw / df_bw
MSE_wi <- SS_wi / df_wi
```

### 2.5. Calculate F statistic

```{r aov_f}
F_stat <- MSE_bw / MSE_wi
F_stat
```

```{r aov_tbl}
anova_table <- data.frame(Df = c(df_bw, df_wi), 
                          Sum_Sq = c(SS_bw, SS_wi),
                          Mean_Sq = c(MSE_bw, MSE_wi),
                          F_stat = c(F_stat, NA))
anova_table
```

## 3. Calculate critical values/p value

### Critical values

```{r aov_crit}
F_crit <- qf(1 - 0.05, df1 = df_bw, df2 = df_wi)

F_stat > F_crit
```

### p value

```{r aov_p_val}
1 - pf(F_stat, df1 = df_bw, df2 = df_wi)
```

## 4. Decide whether to reject/fail to reject $H_0$

-   The calculated test statistic falls within the rejection region

-   p value \< $\alpha$

We reject the null hypothesis.

"With 95% confidence, there is enough evidence to say that at least one of the mean weight changes is significantly different than the others."

"The overall mean weight change was found to be significantly different between diets (ANOVA, p = 0.003)"

### Using `aov()`

```{r aov_r_func}
fit <- aov(Weight_Change~Diet, data = diet_df)
summary(fit)
```

We can further investigate pairwise differences between mean weight change using the Tukey Honest Significant Differences post hoc test:

```{r aov_tukey}
res <- TukeyHSD(fit)
res
```

# Chi-squared Test

Is there an association between gender and diet group? In other words, do gender frequencies differ between diets?

```{r chisq}
table(diet_df$gender, exclude = FALSE)

# replace missing gender values with most frequent
diet_df$gender[is.na(diet_df$gender)] <- 0

# contigency table
tbl <- table(diet_df$Diet, diet_df$gender)
tbl

# relative frequencies
tbl / rowSums(tbl)
barplot(t(tbl / rowSums(tbl)))

chisq.test(tbl)
```

Let's also repeat the example from the slides:

```{r chi2}
vac_tbl <- matrix(c(82, 70, 45, 48, 41, 24, 20, 42), ncol = 2)
colnames(vac_tbl) <- c("Protected", "Not")
rownames(vac_tbl) <- paste0("Vaccine", 1:4)
vac_tbl

# percentages
perc_tbl <- vac_tbl / rowSums(vac_tbl) * 100
barplot(t(perc_tbl))

chisq.test(vac_tbl)

### Post hoc analysis
post_hoc <- c()
for (i in 1:3) {
    for (j in (i + 1):4) {
        v1 <- rownames(vac_tbl)[i]
        v2 <- rownames(vac_tbl)[j]
        
        res <- chisq.test(vac_tbl[c(v1, v2), ])
        post_hoc <- rbind(post_hoc,
                          data.frame(v1 = v1, v2 = v2, p = res$p.value))
    }
}
post_hoc$adj_p <- p.adjust(post_hoc$p, method = "fdr")
post_hoc
```
