---
title: "BB503/BB602 - R Training - Week IX"
author: "Ege Ulgen"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assessing Normality

We'll be using "Normal" dataset for this exercise. This dataset contains 2 continuous variables where one is an example of normally distributed data and the other one is an example of skewed data.

```{r data}
norm_df <- read.csv("../data/Normal_R.csv")

head(norm_df, 3)
```

## Histogram

```{r hist}
par(mfrow = c(1, 2))
hist(norm_df$normal_example, main = "Normal")
hist(norm_df$skewed_example, main = "Skewed")
par(mfrow = c(1, 1))
```

## Quantile-Quantile Plot

```{r qqplot}
# manually
emp_vals <- sort(norm_df$normal_example)
theor_quants <- qnorm((seq_along(emp_vals) - .5) / length(emp_vals))
plot(theor_quants, emp_vals)

# using base R function 'qqnorm' 
par(mfrow = c(1, 2))
qqnorm(norm_df$normal_example)
qqline(norm_df$normal_example)

qqnorm(norm_df$skewed_example)
qqline(norm_df$skewed_example)
par(mfrow = c(1, 1))

# using 'ggqqplot' from 'ggpubr'
# install.packages("ggpubr")
library(ggpubr)

ggqqplot(norm_df, "normal_example")
ggqqplot(norm_df, "skewed_example")
```

## Shapiro--Wilk Test of Normality

$H_0:\ the\ population\ is\ normally\ distributed$

```{r shapiro}
shapiro.test(norm_df$normal_example)
shapiro.test(norm_df$skewed_example)
```

## Kolmogorov-Smirnov Tests

```{r KS0, echo=FALSE}
x <- norm_df$skewed_example
plot(seq(0, 10, .1), pnorm(seq(0, 10, .1), mean = 2, sd = 4), type = "l", lwd = 2, col = "red", ylab = "Cumulative Probability", xlab = "")
lines( sort(x), (1:length(x))/length(x), type="s", lwd = 1.5, col = "blue")

```

```{r KS1}
# does x come from the selected distribution with the specified parameters?
ks.test(norm_df$normal_example, pnorm, mean = mean(norm_df$normal_example), sd = sd(norm_df$normal_example))
```

```{r KS2}
# Do x and y come from the same distribution?
ks.test(norm_df$normal_example, norm_df$skewed_example)
```

# Non-parametric Tests

## Wilcoxon Rank Sum Test

We'll work on the `ToothGrowth` dataset for this exercise. The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received vitamin C by one of two delivery methods, orange juice (OJ) or ascorbic acid (VC). We'll compare the mean lengths between VC and OJ.

```{r wilcox}
?ToothGrowth
head(ToothGrowth, 3)

# Hypothesis: mean tooth lengths of VC and OJ are different
summary(ToothGrowth$len[ToothGrowth$supp == "VC"])
summary(ToothGrowth$len[ToothGrowth$supp == "OJ"])

g <- ggboxplot(data = ToothGrowth,
               x = "supp", y = "len",
               color = "supp",
               xlab = "Supplement", ylab = "Tooth length")
g

# Normal distribution?
shapiro.test(ToothGrowth$len[ToothGrowth$supp == "VC"])
shapiro.test(ToothGrowth$len[ToothGrowth$supp == "OJ"])

# Wilcox test (Mann-Whitney U test)
wilcox.test(len~supp, data = ToothGrowth)

g + stat_compare_means(method = "wilcox")
```

## Kruskal-Wallis Rank Sum Test

We'll use the diet dataset which contains information on 78 people using one of three diets.

```{r kw1}
diet_df <- read.csv("../data/Diet_R.csv")

head(diet_df)

# turn categorical variables into factors
diet_df$Diet <- as.factor(diet_df$Diet)
diet_df$gender <- as.factor(diet_df$gender)

# create new variable
diet_df$Weight_Change <- diet_df$weight6weeks - diet_df$pre.weight


summary(diet_df)

g <- ggboxplot(diet_df, x = "Diet", y = "Weight_Change", color = "Diet")
g
```

We'll compare weight changes of the three diet groups. Let's check the normality of weight changes of the three groups:

```{r kw2}
ggarrange(ggqqplot(diet_df[diet_df$Diet == 1, ], "Weight_Change", title = "Diet 1"),
          ggqqplot(diet_df[diet_df$Diet == 2, ], "Weight_Change", title = "Diet 2"),
          ggqqplot(diet_df[diet_df$Diet == 3, ], "Weight_Change", title = "Diet 3"), ncol = 3)

```

The normality assumption of ANOVA is not met, we'll use the Kruskal-Wallis test instead:

```{r kw3}
kruskal.test(Weight_Change~Diet, data = diet_df)
g + stat_compare_means()

pairwise.wilcox.test(diet_df$Weight_Change, diet_df$Diet)
g + stat_compare_means(comparisons = list(c(1, 2), c(1, 3), c(2, 3)), method = "wilcox")
```

# Multiple Testing Correction

```{r mtc}
# install.packages("multtest")
data(golub, package="multtest")

dim(golub)
row.names(golub) <- paste("gene_", 1:nrow(golub), sep="")
golub[1:3, 1:4]

# we know that groups are 1:27 vs. 28:38
p_plain <- apply(golub, 1, function(x) t.test(x[1:27], x[28:38])$p.value)

### implementations may be slightly different than in the lecture
?p.adjust
p_bonf <- p.adjust(p_plain, method = "bonferroni")
p_holm <- p.adjust(p_plain, method = "holm")
p_fdr <- p.adjust(p_plain, method = "fdr")

head(sort(p_bonf), 10)
head(sort(p_holm), 10)
head(sort(p_fdr), 10)

sum(p_bonf <= 0.05)
sum(p_holm <= 0.05)
sum(p_fdr <= 0.05) # less conservative
```
